The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2011-09-06 projects/slrsim/paper/slrsim.tex
Tiny updates.
Slrsim updates.

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\em Drosophila}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}

\author{{Gregory Jordan and Nick Goldman}\\
\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
%\authorlist{Jordan and Goldman}
%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

%\mbeabstract{% 
\section*{Abstract}

The effects of alignment error on many types of evolutionary inference
are not well understood. Alignment filters are commonly applied to try
to reduce errors resulting from misalignment, but neither the
prevalence of such errors nor the ability of filters to reduce them
has been systematically studied. Focusing on the sitewise detection of
positive selection, we performed simulation experiments to quantify
the false positives and false negatives introduced by alignment error
and the ability of alignment filters to improve performance. We found
that some aligners led to many false positive results across a wide
range of divergence levels, while others resulted in very few. False
negatives were a problem for all aligners, increasing with
divergence. Of the aligners tested, PRANK consistently performed the
best and ClustalW performed the worst. The best alignment filters
showed good ability to reduce the error rates from ClustalW
alignments, but they failed to improve the performance of PRANK
alignments under most circumstances. Of the filters tested, GUIDANCE
performed the best and Gblocks performed the worst, failing to improve
even the error-prone ClustalW alignments. Controls using the true
alignment and an optimal filtering method suggested that performance
improvements could be gained by improving aligners or filters to
reduce the prevalence of false negatives, especially at higher
divergence levels and indel rates. As no combination of aligner plus
filter significantly outperformed the best aligner alone, we conclude
that for the sitewise evolutionary analysis of most vertebrate, insect
and fungal phylogenies, the use of unfiltered PRANK alignments is
recommended.

%}

%\keywords{Alignment, positive selection, alignment filtering, indels}

@@

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach is similar to that of Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. Our experimental design also shares some
features with Fletcher and Yang's \citeyearpar{Fletcher2010Effect}
investigation of the performance of branch-site method for detecting
genewise positive selection under alignment error. Like Fletcher and
Yang \citeyearpar{Fletcher2010Effect}, we simulate codon alignments
with indels and calculate inferred alignments using a small but
diverse sample of aligners. Unlike their study, here we focus on \sw
detection of positive selection occurring throughout a phylogeny (as
opposed to genewise detection of selection acting at specific
branches), we explore a wider range of plausible tree sizes, indel
rates and divergence levels, and we evaluate the impact of a number of
alignment filtering methods on the \sw analysis. Given the recent
proliferation of completed and planned vertebrate genomes, we pay
special attention to choosing simulation parameters and analysis
methods similar to those commonly encountered in the \sw analysis of
vertebrate gene families.

@@

GUIDANCE and T-Coffee produce individual scores for each residue,
allowing individual residues to be masked instead of entire
columns. GUIDANCE generates many replicate alignments, each using a
slightly perturbed guide tree, with either MAFFT or \pranka as the
bootstrap aligner. The program then assigns to each residue from the
input alignment a score from 0 to 1 based on how consistently it was
placed in the replicate alignments. In order to maximize the
similarity between the input aligner and the bootstrap aligner, we ran
GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
and with 30 \pranka replicates when filtering \prankc
alignments. T-Coffee calculates the residue-wise consistency between
an input multiple alignment and independently calculated pairwise
alignments \citep{Notredame2003Using}, rounding and normalizing
residue scores into integers between 0 and 9. T-Coffee was run using
its default settings and the {\em evaluate\_mode
  --output=score\_ascii} command-line parameters to output alignment
scores.

@@

\begin{figure*}[t]
\begin{center}
%\includegraphics[scale=0.8]{fig4.pdf}
\includegraphics[scale=0.8]{fig4.pdf}
\end{center}
\caption{Sequences were simulated with indels using one of three tree
  shapes (t-, 17- or 44-taxon) and a range of indel rates and mean
  path length (MPL) divergence levels. Alignments were inferred with
  one of four aligners (ClustalW, MAFFT, \pranka{}, \prankc) and
  analyzed with SLR; true alignments were separately analyzed with
  SLR. One hundred replicates were simulated for each set of
  conditions. Each cell is colored according to the performance at a
  given (indel rate, MPL) pair as measured by one of three summary
  statistics: (a) the true positive rate (TPR) at the recommended
  cutoff threshold, (b) the false positive rate (FPR) at the
  recommended cutoff threshold, or (c) the TPR at a 1\% FPR
  threshold. Results for MAFFT and \pranka are omitted from (a) and
  (b); as in (c) they show characteristics intermediate between
  ClustalW and \prankc.}
\label{fig_4}
\end{figure*}

@@

\bibliographystyle{mbe3} \fontsize{9}{10}\selectfont% \itemsep 0pt

\bibliography{slrsim}

>>>>

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}
%\usepackage{ulem}

@@

\renewcommand\refname{Literature Cited}

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in mammals}
%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrate genomics}
\author{{Gregory Jordan and Nick Goldman}}
%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\maketitle
%\authorlist{Jordan and Goldman}
%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

%\mbeabstract{% 
\section*{Abstract}
\textcolor{red}{
Alignment filters are commonly applied to try to reduce errors
resulting from misalignment, but the prevalence of such errors and the
ability of filters to reduce them is not well understood.
% 
Focusing on the sitewise detection of positive selection across a wide
range of divergence levels and indel rates, we performed simulation
experiments to quantify the false positives and false negatives
introduced by alignment error and the ability of alignment filters to
improve performance.
%
We found that some aligners led to many false positives while others
resulted in very few.  False negatives were a problem for all
aligners, increasing with divergence. Of the aligners tested, PRANK's
codon-based alignments consistently performed the best and ClustalW
performed the worst.
%
Of the filters tested, GUIDANCE performed the best and Gblocks
performed the worst. Although some filters showed good ability to
reduce the error rates from ClustalW and MAFFT alignments, none were
found to substantially improve the performance of PRANK alignments
under most \textcolor{red}{conditions}.
% 
Our results revealed distinct trends in error rates and power levels
for aligners and filters within a biologically plausible parameter
space. Controls using the true alignment and an optimal filtering
method suggested that performance improvements could be gained by
improving aligners or filters to reduce the prevalence of false
negatives, especially at higher divergence levels and indel rates.
}
%\keywords{Alignment, positive selection, alignment filtering, indels}

@@

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach builds from that of Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. \textcolor{red}{As we simulate codon
  alignments with indels and calculate inferred alignments using a
  small but diverse sample of aligners and alignment filters, the
  experimental design also shares some features with Fletcher and
  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
  performance of the branch-site method for detecting positive
  selection under alignment error and Privman et al.'s
  \citeyearpar{Privman2011Improving} study of the effect of alignment
  filtering on detecting positive selection in HIV-1 genes. However,
  the present work differs from previous studies in a number of
  important ways.}

\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
  detection of sitewise positive selection occurring throughout a
  phylogeny (as opposed to genewise detection of selection acting at
  specific branches) and we evaluate the impact of a number of
  alignment filtering methods on the \sw analysis.}

\textcolor{red}{Privman et al., whose simulations centered on the
  evolutionary chracteristics of three HIV-1 genes, concluded that
  alignment filtering improves the performance of positive selection
  inference by removing false positives. While this may be true for
  the three genes (and perhaps for HIV-1 in general), HIV-1 is known
  to evolve with widespread positive selection in the human host
  \citep{Yang2003Widespread}. Results valid for these three HIV-1
  genes may not be widely applicable to large-scale vertebrate and
  mammalian comparative datasets, which exhibit less adaptive
  evolution \citep{Kosiol2008Patterns} and which comprise a larger
  diversity of protein structures and a wider range of species
  divergence levels.}

\textcolor{red}{We hypothesize that these factors may strongly affect
  the the performance of methods for alignment and detection of
  selection. Accordingly, our simulations employ parameters similar to
  those commonly encountered in the \sw analysis of vertebrate gene
  families and encompass a wide range of biologically plausible indel
  rates and divergence levels. Although our main purpose is to
  understand the behaviour of various aligners and filters within our
  range of simulations, we will make use of the above-noted
  methodological similarities by pointing out those aspects of our
  study which agree with or contradict the literature where possible.}

@@

GUIDANCE and T-Coffee produce individual scores for each residue,
allowing individual residues to be masked instead of entire
columns. \textcolor{red}{Privman et al. \citeyearpar{Privman2011Improving} found a
residue-based filter to be more effective than its column-based
equivalent, so we filtered residues instead of alignment columns where
possible}. GUIDANCE generates many replicate alignments, each using a
slightly perturbed guide tree, with either MAFFT or \pranka as the
bootstrap aligner. The program then assigns to each residue from the
input alignment a score from 0 to 1 based on how consistently it was
placed in the replicate alignments. In order to maximize the
similarity between the input aligner and the bootstrap aligner, we ran
GUIDANCE with 100 MAFFT replicates when filtering ClustalW alignments
and with 30 \pranka replicates when filtering \prankc
alignments. T-Coffee calculates the residue-wise consistency between
an input multiple alignment and independently calculated pairwise
alignments \citep{Notredame2003Using}, rounding and normalizing
residue scores into integers between 0 and 9. T-Coffee was run using
its default settings and the {\em evaluate\_mode
  --output=score\_ascii} command-line parameters to output alignment
scores.

@@

\begin{figure*}[t]
\begin{center}
%\includegraphics[scale=0.8]{fig4.pdf}
\includegraphics[scale=0.8]{fig4.pdf}
\end{center}
\caption{Sequences were simulated with indels using one of three tree
  shapes (6-, 17- or 44-taxon) and a range of indel rates and mean
  path length (MPL) divergence levels. Alignments were inferred with
  one of four aligners (ClustalW, MAFFT, \pranka{}, \prankc) and
  analyzed with SLR; true alignments were separately analyzed with
  SLR. One hundred replicates were simulated for each set of
  conditions. Each cell is colored according to the performance at a
  given (indel rate, MPL) pair as measured by one of three summary
  statistics: (a) the true positive rate (TPR) at the recommended
  cutoff threshold, (b) the false positive rate (FPR) at the
  recommended cutoff threshold, or (c) the TPR at a 1\% FPR
  threshold. Results for MAFFT and \pranka are omitted from (a) and
  (b); as in (c) they show characteristics intermediate between
  ClustalW and \prankc.}
\label{fig_4}
\end{figure*}

@@

\bibliographystyle{mbe3} \fontsize{9}{10}\selectfont% \itemsep 0pt
\bibliography{slrsim}
